
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.width = 10, fig.height = 6, dpi = 150
)
```

```{r packages}
# Install if needed

library(tidyverse)
library(lme4)
library(lmerTest)
library(emmeans)
library(ordinal)
library(kableExtra)
library(broom.mixed)
library(patchwork)
library(scales)
```

```{r}
# --- Llama results ---
traits_llama  <- read_csv("llama_Llama-3.1-8B-Instruct_traits_raw.csv")
guilt_llama   <- read_csv("llama_Llama-3.1-8B-Instruct_guilt_raw.csv")
occup_llama   <- read_csv("llama_Llama-3.1-8B-Instruct_occupations_raw.csv")

# --- Claude results ---
traits_claude <- read_csv("claude_claude-sonnet-4-6_traits_raw.csv")
guilt_claude  <- read_csv("claude_claude-sonnet-4-6_guilt_raw.csv")
occup_claude  <- read_csv("claude_claude-sonnet-4-6_occupations_raw.csv")

# Combine per experiment
traits_all <- bind_rows(traits_llama, traits_claude)
guilt_all  <- bind_rows(guilt_llama, guilt_claude)
occup_all  <- bind_rows(occup_llama, occup_claude)

cat("Trait rows:", nrow(traits_all),
    "| Guilt rows:", nrow(guilt_all),
    "| Occupation rows:", nrow(occup_all), "\n")
```


#Data Preparation

```{r}
prep_factors <- function(df) {
  df %>%
    mutate(
      dialect    = factor(dialect, levels = c("aae", "sae")),
      oversight  = factor(oversight, levels = c("no_training", "training")),
      scratchpad = factor(scratchpad, levels = c("no_scratchpad", "with_scratchpad")),
      model      = factor(model),
      stimulus_id = factor(stimulus_id)
    )
}

traits_all <- prep_factors(traits_all)
guilt_all  <- prep_factors(guilt_all)
occup_all  <- prep_factors(occup_all)
```

## Trait Index Computation

```{r}
pos_traits <- c("trait_alert", "trait_ambitious", "trait_artistic", "trait_brilliant",
                "trait_efficient", "trait_faithful", "trait_generous", "trait_honest",
                "trait_imaginative", "trait_intelligent", "trait_kind", "trait_loyal",
                "trait_musical", "trait_neat", "trait_passionate", "trait_persistent",
                "trait_practical", "trait_progressive", "trait_quiet", "trait_sensitive",
                "trait_sophisticated", "trait_straightforward")

neg_traits <- c("trait_aggressive", "trait_conservative", "trait_conventional",
                "trait_cruel", "trait_dirty", "trait_ignorant", "trait_lazy",
                "trait_loud", "trait_radical", "trait_reserved", "trait_rude",
                "trait_stubborn", "trait_stupid", "trait_suspicious")

traits_all <- traits_all %>%
  mutate(
    across(all_of(neg_traits), ~ 11 - .x, .names = "{.col}_corrected"),
    across(all_of(pos_traits), ~ .x, .names = "{.col}_corrected")
  )

corrected_cols <- c(paste0(pos_traits, "_corrected"), paste0(neg_traits, "_corrected"))

traits_all <- traits_all %>%
  rowwise() %>%
  mutate(
    trait_score_unified  = mean(c_across(all_of(corrected_cols)), na.rm = TRUE),
    trait_score_positive = mean(c_across(all_of(pos_traits)), na.rm = TRUE),
    trait_score_negative = mean(c_across(all_of(neg_traits)), na.rm = TRUE)
  ) %>%
  ungroup()
```

## Refusal Coding

```{r}
# Code refusal as binary: 1 = refused (all traits NA), 0 = complied
traits_all <- traits_all %>%
  mutate(
    refused = as.integer(is.na(trait_alert))
  )

traits_all %>%
  group_by(model) %>%
  summarise(
    total = n(),
    refused = sum(refused, na.rm = TRUE),
    complied = sum(!refused, na.rm = TRUE),
    refusal_rate = mean(refused, na.rm = TRUE)
  ) %>%
  kable(digits = 3, caption = "Refusal rates by model") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## Parse Success Rates

```{r}
tibble(
  Experiment = c("Traits (Llama)", "Traits (Claude)", "Traits (Claude) - compliance only",
                 "Guilt", "Occupations"),
  `Total rows` = c(
    nrow(traits_llama), nrow(traits_claude),
    sum(!traits_all$refused[traits_all$model == unique(traits_claude$model)[1]], na.rm = TRUE),
    nrow(guilt_all), nrow(occup_all)),
  `Parse success` = c(
    mean(!is.na(traits_llama$trait_alert)),
    mean(!is.na(traits_claude$trait_alert)),
    mean(!is.na(filter(traits_all, model == unique(traits_claude$model)[1], !refused)$trait_alert)),
    mean(!is.na(guilt_all$guilty)),
    mean(!is.na(occup_all$prestige))
  )
) %>%
  mutate(`Parse success` = scales::percent(`Parse success`, accuracy = 0.1)) %>%
  kable(caption = "Parse success rates") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## Refusal Rates by Condition

```{r}
claude_refusal <- traits_all %>%
  filter(str_detect(model, "claude"))

claude_refusal %>%
  group_by(dialect, oversight, scratchpad) %>%
  summarise(
    n = n(),
    n_refused = sum(refused),
    refusal_rate = mean(refused),
    .groups = "drop"
  ) %>%
  kable(digits = 3, caption = "Claude refusal rates by condition") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## Refusal Rate Visualisation

```{r}
claude_refusal %>%
  group_by(dialect, oversight, scratchpad) %>%
  summarise(refusal_rate = mean(refused),
            se = sqrt(refusal_rate * (1 - refusal_rate) / n()),
            .groups = "drop") %>%
  ggplot(aes(x = interaction(oversight, scratchpad, sep = "\n"),
             y = refusal_rate, fill = dialect)) +
  geom_col(position = position_dodge(0.8), width = 0.7) +
  geom_errorbar(aes(ymin = refusal_rate - se, ymax = refusal_rate + se),
                position = position_dodge(0.8), width = 0.25) +
  labs(title = "Claude: Selective Refusal Rate by Condition",
       subtitle = "Claude refuses to rate AAE speakers 2-3x more often than SAE speakers",
       x = "Condition (Oversight / Scratchpad)",
       y = "Proportion Refused",
       fill = "Dialect") +
  scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(labels = scales::percent, limits = c(0, 0.8)) +
  theme_minimal(base_size = 13)
```

## Logistic Mixed-Effects Model for Refusal

```{r}
refusal_model <- glmer(
  refused ~ dialect * oversight * scratchpad + (1 | stimulus_id),
  data = claude_refusal,
  family = binomial,
  control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 100000))
)

summary(refusal_model)
```

```{r}
car::Anova(refusal_model, type = "III") %>%
  as.data.frame() %>%
  rownames_to_column("Effect") %>%
  mutate(sig = case_when(`Pr(>Chisq)` < 0.001 ~ "***",
                         `Pr(>Chisq)` < 0.01  ~ "**",
                         `Pr(>Chisq)` < 0.05  ~ "*",
                         TRUE ~ "")) %>%
  kable(digits = 4, caption = "Type III Wald tests: Claude refusal") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

```{r}
emmeans(refusal_model, pairwise ~ dialect, type = "response")
```

# Trait Evaluation Analysis

## Descriptive Statistics

```{r}
traits_all %>%
  filter(!is.na(trait_score_unified)) %>%
  group_by(model, dialect, oversight, scratchpad) %>%
  summarise(
    mean_unified  = mean(trait_score_unified, na.rm = TRUE),
    sd_unified    = sd(trait_score_unified, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) %>%
  kable(digits = 3, caption = "Trait scores by condition (parsed trials only)") %>%
  kable_styling(bootstrap_options = "striped", font_size = 12)
```

## Mixed-Effects Models

### Llama

```{r}
trait_llama_data <- traits_all %>%
  filter(str_detect(model, "llama"), !is.na(trait_score_unified))

trait_model_llama <- lmer(
  trait_score_unified ~ dialect * oversight * scratchpad + (1 | stimulus_id),
  data = trait_llama_data
)

summary(trait_model_llama)
```

```{r}
anova(trait_model_llama) %>%
  as.data.frame() %>%
  rownames_to_column("Effect") %>%
  mutate(sig = case_when(`Pr(>F)` < 0.001 ~ "***",
                         `Pr(>F)` < 0.01  ~ "**",
                         `Pr(>F)` < 0.05  ~ "*",
                         TRUE ~ "")) %>%
  kable(digits = 4, caption = "ANOVA: Trait scores (Llama)") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

```{r}
emmeans(trait_model_llama, pairwise ~ dialect)
```

### Claude (complied trials only)

```{r}
trait_claude_data <- traits_all %>%
  filter(str_detect(model, "claude"), !refused, !is.na(trait_score_unified))

cat("Claude complied trials with valid scores:", nrow(trait_claude_data), "\n")

trait_model_claude <- lmer(
  trait_score_unified ~ dialect * oversight * scratchpad + (1 | stimulus_id),
  data = trait_claude_data
)

summary(trait_model_claude)
```

```{r trait-claude-anova}
anova(trait_model_claude) %>%
  as.data.frame() %>%
  rownames_to_column("Effect") %>%
  mutate(sig = case_when(`Pr(>F)` < 0.001 ~ "***",
                         `Pr(>F)` < 0.01  ~ "**",
                         `Pr(>F)` < 0.05  ~ "*",
                         TRUE ~ "")) %>%
  kable(digits = 4, caption = "ANOVA: Trait scores (Claude, complied only)") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## Valence-Specific Analyses

```{r}
pos_model_llama <- lmer(
  trait_score_positive ~ dialect * oversight * scratchpad + (1 | stimulus_id),
  data = trait_llama_data
)

neg_model_llama <- lmer(
  trait_score_negative ~ dialect * oversight * scratchpad + (1 | stimulus_id),
  data = trait_llama_data
)

anova(pos_model_llama)

anova(neg_model_llama)
```

---

# Guilt Attribution Analysis

## Descriptive Statistics

```{r}
guilt_all %>%
  filter(!is.na(guilty)) %>%
  group_by(model, dialect, oversight, scratchpad) %>%
  summarise(
    guilt_rate = mean(guilty),
    n_guilty   = sum(guilty),
    n_total    = n(),
    .groups = "drop"
  ) %>%
  kable(digits = 3, caption = "Guilt rates by condition") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

```{r}
guilt_all %>%
  filter(!is.na(guilty)) %>%
  group_by(model, dialect, oversight, scratchpad) %>%
  summarise(
    n_guilty = sum(guilty),
    n_total  = n(),
    cell     = sprintf("%d/%d (%.1f%%)", n_guilty, n_total, 100 * n_guilty / n_total),
    .groups  = "drop"
  ) %>%
  mutate(
    condition = paste(scratchpad, oversight, sep = " / "),
    model = if_else(str_detect(model, "claude"), "Claude", "Llama")
  ) %>%
  select(model, dialect, condition, cell) %>%
  pivot_wider(names_from = condition, values_from = cell) %>%
  arrange(model, dialect) %>%
  kable(caption = "Guilty verdicts by model and condition (n guilty / n total)") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

```{r}
guilt_all %>%
  filter(!is.na(guilty)) %>%
  group_by(model) %>%
  summarise(
    total_guilty = sum(guilty),
    total_trials = n(),
    rate = mean(guilty),
    .groups = "drop"
  ) %>%
  kable(digits = 4, caption = "Total guilty verdicts by model") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## Logistic Mixed-Effects Models

### Llama

```{r}
guilt_llama_data <- guilt_all %>%
  filter(str_detect(model, "llama"), !is.na(guilty))

guilt_model_llama <- glmer(
  guilty ~ dialect * oversight * scratchpad + (1 | stimulus_id),
  data = guilt_llama_data,
  family = binomial,
  control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 100000))
)

summary(guilt_model_llama)
```

```{r}
car::Anova(guilt_model_llama, type = "III") %>%
  as.data.frame() %>%
  rownames_to_column("Effect") %>%
  mutate(sig = case_when(`Pr(>Chisq)` < 0.001 ~ "***",
                         `Pr(>Chisq)` < 0.01  ~ "**",
                         `Pr(>Chisq)` < 0.05  ~ "*",
                         TRUE ~ "")) %>%
  kable(digits = 4, caption = "Type III Wald tests: Guilt (Llama)") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

### Claude

```{r}
guilt_claude_data <- guilt_all %>%
  filter(str_detect(model, "claude"), !is.na(guilty))

cat("Claude guilt: ", sum(guilt_claude_data$guilty), "guilty out of",
    nrow(guilt_claude_data), "trials\n")
cat("All guilty verdicts AAE?",
    all(guilt_claude_data$guilty[guilt_claude_data$dialect == "sae"] == 0), "\n\n")

guilt_model_claude <- glmer(
    guilty ~ dialect * oversight * scratchpad + (1 | stimulus_id),
    data = guilt_claude_data,
    family = binomial,
    control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 100000))
  )
summary(guilt_model_claude)
```

---

# Occupational Assignment Analysis

## Descriptive Statistics

```{r}
occup_all %>%
  filter(!is.na(prestige)) %>%
  group_by(model, dialect, oversight, scratchpad) %>%
  summarise(
    mean_prestige = mean(prestige),
    sd_prestige   = sd(prestige),
    n = n(),
    .groups = "drop"
  ) %>%
  kable(digits = 3, caption = "Occupational prestige by condition") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## Top Occupations by Dialect and Model

```{r}
occup_all %>%
  filter(!is.na(occupation)) %>%
  count(model, dialect, occupation, sort = TRUE) %>%
  group_by(model, dialect) %>%
  slice_head(n = 10) %>%
  pivot_wider(names_from = dialect, values_from = n, values_fill = 0) %>%
  kable(caption = "Top 10 occupations by dialect and model") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## Stereotype-Associated Occupations

```{r}
occup_all %>%
  filter(!is.na(occupation)) %>%
  count(model, dialect, occupation) %>%
  pivot_wider(names_from = dialect, values_from = n, values_fill = 0) %>%
  mutate(
    ratio = ifelse(sae > 0, aae / sae, Inf),
    diff = aae - sae,
    total = aae + sae
  ) %>%
  filter(total >= 10) %>%
  arrange(desc(abs(diff))) %>%
  group_by(model) %>%
  slice_head(n = 10) %>%
  kable(digits = 2, caption = "Occupations with largest dialect asymmetry (min 10 total)") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## Visualisation

```{r}
occup_all %>%
  filter(!is.na(prestige)) %>%
  group_by(model, dialect, oversight, scratchpad) %>%
  summarise(mean_prestige = mean(prestige),
            se = sd(prestige) / sqrt(n()),
            .groups = "drop") %>%
  ggplot(aes(x = interaction(oversight, scratchpad, sep = "\n"),
             y = mean_prestige, fill = dialect)) +
  geom_col(position = position_dodge(0.8), width = 0.7) +
  geom_errorbar(aes(ymin = mean_prestige - se, ymax = mean_prestige + se),
                position = position_dodge(0.8), width = 0.25) +
  facet_wrap(~ model) +
  labs(title = "Mean Occupational Prestige by Condition",
       x = "Condition (Oversight / Scratchpad)",
       y = "Mean Prestige (0-2)",
       fill = "Dialect") +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal(base_size = 13)
```

## Ordinal Mixed-Effects Models

```{r}
occup_llama_data <- occup_all %>%
  filter(str_detect(model, "llama"), !is.na(prestige)) %>%
  mutate(prestige_ord = ordered(prestige))

occup_model_llama <- clmm(
  prestige_ord ~ dialect * oversight * scratchpad + (1 | stimulus_id),
  data = occup_llama_data
)

summary(occup_model_llama)
```

```{r}
occup_claude_data <- occup_all %>%
  filter(str_detect(model, "claude"), !is.na(prestige)) %>%
  mutate(prestige_ord = ordered(prestige))

occup_model_claude <- clmm(
  prestige_ord ~ dialect * oversight * scratchpad + (1 | stimulus_id),
  data = occup_claude_data
)

summary(occup_model_claude)
```

## Within-Stimulus Occupation Mismatch Analysis

```{r}
pair_occupations <- function(df, model_label) {
  aae <- df %>%
    filter(dialect == "aae") %>%
    select(stimulus_id, oversight, scratchpad, rep, occupation, prestige)
  names(aae)[5:6] <- c("occupation_aae", "prestige_aae")

  sae <- df %>%
    filter(dialect == "sae") %>%
    select(stimulus_id, oversight, scratchpad, rep, occupation, prestige)
  names(sae)[5:6] <- c("occupation_sae", "prestige_sae")

  paired <- merge(aae, sae, by = c("stimulus_id", "oversight", "scratchpad", "rep"))
  paired <- paired[!is.na(paired$occupation_aae) & !is.na(paired$occupation_sae), ]
  paired$model <- model_label
  paired$mismatch <- paired$occupation_aae != paired$occupation_sae
  paired$prestige_diff <- paired$prestige_sae - paired$prestige_aae
  paired$direction <- ifelse(paired$prestige_diff > 0, "SAE higher",
                      ifelse(paired$prestige_diff < 0, "AAE higher", "Equal"))
  paired$pattern <- paste0(paired$occupation_aae, " -> ", paired$occupation_sae)
  paired
}

paired_llama  <- pair_occupations(occup_all %>% filter(str_detect(model, "llama")),
                                  "Llama")
paired_claude <- pair_occupations(occup_all %>% filter(str_detect(model, "claude")),
                                  "Claude")
paired_all <- bind_rows(paired_llama, paired_claude)

cat("Total paired trials: llama =", nrow(paired_llama),
    ", claude =", nrow(paired_claude), "\n")
```

### Mismatch Rates

```{r}
paired_all %>%
  group_by(model) %>%
  summarise(
    total_pairs = n(),
    n_mismatch = sum(mismatch),
    mismatch_rate = mean(mismatch),
    sae_higher = mean(prestige_diff > 0),
    aae_higher = mean(prestige_diff < 0),
    equal_prestige = mean(prestige_diff == 0),
    .groups = "drop"
  ) %>%
  mutate(across(c(mismatch_rate, sae_higher, aae_higher, equal_prestige),
                ~ scales::percent(.x, accuracy = 0.1))) %>%
  kable(caption = "Within-stimulus occupation mismatch summary") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

### Top Mismatch Patterns (Table 4.6)

```{r}
paired_all %>%
  filter(mismatch, prestige_diff != 0) %>%
  count(model, pattern, prestige_diff, sort = TRUE) %>%
  mutate(direction = ifelse(prestige_diff > 0, "up SAE higher", "down AAE higher")) %>%
  group_by(model) %>%
  slice_head(n = 10) %>%
  select(model, `AAE -> SAE` = pattern, n, direction) %>%
  kable(caption = "Most frequent prestige-changing mismatch patterns (same sentence, different occupation)") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

```{r}
splits <- paired_all %>%
  filter(mismatch) %>%
  filter(
    (occupation_aae == "cook" & occupation_sae == "chef") |
    (occupation_aae == "chef" & occupation_sae == "cook") |
    (occupation_aae == "nurse" & occupation_sae == "doctor") |
    (occupation_aae == "doctor" & occupation_sae == "nurse") |
    (occupation_aae == "driver" & occupation_sae == "teacher") |
    (occupation_aae == "teacher" & occupation_sae == "driver")
  ) %>%
  count(model, pattern, prestige_diff) %>%
  mutate(direction = ifelse(prestige_diff > 0, "SAE up", "AAE up"))

splits %>%
  select(model, `AAE -> SAE` = pattern, n, direction) %>%
  kable(caption = "Stereotype-consistent occupation splits: same sentence, different prestige") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

# Cross-Model Comparison

```{r}
cross_model <- bind_rows(
  # Trait bias
  traits_all %>%
    filter(!is.na(trait_score_unified)) %>%
    group_by(model, dialect) %>%
    summarise(value = mean(trait_score_unified), .groups = "drop") %>%
    pivot_wider(names_from = dialect, values_from = value) %>%
    mutate(metric = "Trait Score\n(SAE - AAE)", effect = sae - aae) %>%
    select(model, metric, effect),

  # Guilt bias
  guilt_all %>%
    filter(!is.na(guilty)) %>%
    group_by(model, dialect) %>%
    summarise(value = mean(guilty), .groups = "drop") %>%
    pivot_wider(names_from = dialect, values_from = value) %>%
    mutate(metric = "Guilt Rate\n(AAE - SAE)", effect = aae - sae) %>%
    select(model, metric, effect),

  # Occupation bias
  occup_all %>%
    filter(!is.na(prestige)) %>%
    group_by(model, dialect) %>%
    summarise(value = mean(prestige), .groups = "drop") %>%
    pivot_wider(names_from = dialect, values_from = value) %>%
    mutate(metric = "Prestige\n(SAE - AAE)", effect = sae - aae) %>%
    select(model, metric, effect),

  # Refusal bias (Claude)
  traits_all %>%
    filter(str_detect(model, "claude")) %>%
    group_by(model, dialect) %>%
    summarise(value = mean(refused), .groups = "drop") %>%
    pivot_wider(names_from = dialect, values_from = value) %>%
    mutate(metric = "Refusal Rate\n(AAE - SAE)", effect = aae - sae) %>%
    select(model, metric, effect)
)

ggplot(cross_model, aes(x = metric, y = effect, fill = model)) +
  geom_col(position = position_dodge(0.7), width = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  labs(title = "Dialect Bias Across Tasks and Models",
       subtitle = "Positive values = bias against AAE speakers",
       x = "", y = "Effect (dialect difference)", fill = "Model") +
  scale_fill_brewer(palette = "Set1") +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(size = 11))
```

---

# Hypothesis Testing Summary

```{r}
get_trait_pvals <- function(mod, model_label) {
  trait_anova <- anova(mod) %>% as.data.frame() %>% rownames_to_column("Effect")

  p_h1 <- trait_anova %>% filter(Effect == "dialect") %>% pull(`Pr(>F)`)
  p_h2 <- trait_anova %>% filter(Effect == "dialect:oversight") %>% pull(`Pr(>F)`)
  p_h3 <- trait_anova %>% filter(Effect == "dialect:scratchpad") %>% pull(`Pr(>F)`)

  raw_p <- c(H1 = p_h1, H2 = p_h2, H3 = p_h3)
  adj_p <- p.adjust(raw_p, method = "holm")

  tibble(
    Model = model_label,
    Hypothesis = c("H1: Dialect main effect",
                   "H2: Dialect x Oversight",
                   "H3: Dialect x Scratchpad"),
    `Raw p` = raw_p,
    `Holm-adjusted p` = adj_p,
    Significant = adj_p < 0.05
  )
}

tryCatch({
  summary_table <- bind_rows(
    get_trait_pvals(trait_model_llama, "Llama (traits)"),
    get_trait_pvals(trait_model_claude, "Claude (traits, complied)")
  )

  ref_anova <- car::Anova(refusal_model, type = "III") %>%
    as.data.frame() %>%
    rownames_to_column("Effect")

  ref_pvals <- tibble(
    Model = "Claude (refusal)",
    Hypothesis = c("Dialect effect on refusal",
                   "Dialect x Oversight on refusal",
                   "Dialect x Scratchpad on refusal"),
    `Raw p` = c(
      ref_anova %>% filter(Effect == "dialect") %>% pull(`Pr(>Chisq)`),
      ref_anova %>% filter(Effect == "dialect:oversight") %>% pull(`Pr(>Chisq)`),
      ref_anova %>% filter(Effect == "dialect:scratchpad") %>% pull(`Pr(>Chisq)`)
    ),
    `Holm-adjusted p` = p.adjust(`Raw p`, method = "holm"),
    Significant = `Holm-adjusted p` < 0.05
  )

  bind_rows(summary_table, ref_pvals) %>%
    kable(digits = 4, caption = "Hypothesis tests with Holm correction") %>%
    kable_styling(bootstrap_options = "striped", full_width = FALSE)
}, error = function(e) {
  cat("Could not generate summary:", conditionMessage(e), "\n")
})
```

---

# Scratchpad Content Analysis

```{r}
scratchpad_data <- traits_all %>%
  filter(scratchpad == "with_scratchpad",
         !is.na(scratchpad_text), scratchpad_text != "")

cat("Trials with scratchpad text:", nrow(scratchpad_data), "\n")

scratchpad_data %>%
  group_by(model, dialect) %>%
  summarise(
    n = n(),
    mean_score = mean(trait_score_unified, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  kable(digits = 3, caption = "Scratchpad trials by model and dialect") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

```{r scratchpad-keywords}
neg_stereotype <- c("uneducated", "informal", "slang", "street", "urban",
                    "unprofessional", "improper", "incorrect", "broken",
                    "nonstandard", "aave", "vernacular", "dialect")

pos_stereotype <- c("educated", "professional", "formal", "proper",
                    "articulate", "eloquent", "standard", "correct",
                    "well-spoken", "grammatical")

scratchpad_keywords <- scratchpad_data %>%
  mutate(text_lower = tolower(scratchpad_text)) %>%
  rowwise() %>%
  mutate(
    neg_terms = sum(str_count(text_lower, neg_stereotype)),
    pos_terms = sum(str_count(text_lower, pos_stereotype)),
    dialect_mention = str_detect(text_lower, "aave|african american|vernacular|dialect|slang")
  ) %>%
  ungroup()

scratchpad_keywords %>%
  group_by(model, dialect) %>%
  summarise(
    mean_neg = mean(neg_terms),
    mean_pos = mean(pos_terms),
    pct_dialect_mention = mean(dialect_mention),
    n = n(),
    .groups = "drop"
  ) %>%
  kable(digits = 3,
        caption = "Stereotype keyword counts and dialect recognition in scratchpad (Table 4.7)") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

# Effect Size Summary

```{r}
cohens_d <- function(data, model_filter, outcome = "trait_score_unified") {
  d <- data %>%
    filter(str_detect(model, model_filter), !is.na(.data[[outcome]]))
  aae <- d %>% filter(dialect == "aae") %>% pull(!!sym(outcome))
  sae <- d %>% filter(dialect == "sae") %>% pull(!!sym(outcome))
  pooled_sd <- sqrt((sd(aae)^2 + sd(sae)^2) / 2)
  d_val <- (mean(sae) - mean(aae)) / pooled_sd
  tibble(Model = model_filter, Outcome = outcome,
         `Cohen's d` = d_val, `Mean AAE` = mean(aae), `Mean SAE` = mean(sae),
         `Pooled SD` = pooled_sd, `N AAE` = length(aae), `N SAE` = length(sae))
}

bind_rows(
  cohens_d(traits_all %>% filter(!refused | is.na(refused)), "llama"),
  cohens_d(traits_all %>% filter(!refused), "claude")
) %>%
  kable(digits = 4, caption = "Effect sizes: Dialect effect on trait scores") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## Figure 4.1: Claude Refusal Rates

```{r}
fig_refusal <- claude_refusal %>%
  mutate(
    scratchpad_label = ifelse(scratchpad == "with_scratchpad", "With Scratchpad", "No Scratchpad"),
    oversight_label  = ifelse(oversight == "training", "Training", "No Training")
  ) %>%
  group_by(dialect, oversight_label, scratchpad_label) %>%
  summarise(refusal_rate = mean(refused),
            se = sqrt(refusal_rate * (1 - refusal_rate) / n()),
            .groups = "drop") %>%
  ggplot(aes(x = oversight_label, y = refusal_rate, fill = dialect)) +
  geom_col(position = position_dodge(0.75), width = 0.65, alpha = 0.9) +
  geom_errorbar(aes(ymin = refusal_rate - se, ymax = refusal_rate + se),
                position = position_dodge(0.75), width = 0.2, linewidth = 0.4) +
  facet_wrap(~ scratchpad_label) +
  scale_fill_manual(values = dialect_colors, labels = dialect_labels, name = "Dialect") +
  scale_y_continuous(labels = percent, limits = c(0, 0.80), breaks = seq(0, 0.8, 0.2)) +
  labs(title = "Claude's Refusal rate on the trait evaluation task by dialect and conditions",
       x = "Oversight Condition", y = "Refusal Rate") +
  theme_thesis

fig_refusal
```

## Figure 4.2: Cross-Model Comparison

```{r}
summary_data <- tribble(
  ~model,   ~metric,                              ~value,
  "Llama",  "Trait bias\n(Cohen's d)",             0.18,
  "Claude", "Trait bias\n(Cohen's d)",            -0.07,
  "Llama",  "Refusal rate\ndifference",            0.00,
  "Claude", "Refusal rate\ndifference",            0.39,
  "Llama",  "Prestige\n(SAE - AAE)",               0.017,
  "Claude", "Prestige\n(SAE - AAE)",               0.026,
)

fig_crossmodel <- ggplot(summary_data, aes(x = metric, y = value, fill = model)) +
  geom_col(position = position_dodge(0.7), width = 0.6, alpha = 0.9) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.4) +
  geom_text(aes(label = sprintf("%.3f", value),
                y = ifelse(value >= 0, value + 0.012, value - 0.012)),
            position = position_dodge(0.7), size = 3.5, family = "serif") +
  scale_fill_manual(values = model_colors, name = "Model") +
  labs(title = "Cross-Model Comparison of Dialect Bias",
       subtitle = "Positive values indicate bias against AAE speakers",
       x = "", y = "Effect size / Rate difference") +
  theme_thesis +
  theme(axis.text.x = element_text(size = 10, lineheight = 1.1))

fig_crossmodel
```

```{r}
ggsave("fig4-1_refusal_rates.png", fig_refusal,   width = 10, height = 5.5, dpi = 300)
ggsave("fig4-2_cross_model.png",   fig_crossmodel, width = 11, height = 5.5, dpi = 300)
```
